The increasing cost of happiness

### 2021-06-12 src/draft-SSM1.Rmd
Minor updates to intro and discusssion

- We investigate how the relationship between income and subjective wellbeing (affective and cognitive) has changed over time, using a dataset with highly detailed measurements of each
- Nineteen years of data from a well-developed, economically stable G20 nation is examined (Australia), allowing a relatively undisturbed view of the changes without major economic shocks
- We are able to show for the first time that the income level at which affective wellbeing is less determined by income has risen since 2001, resulting in greater inequality in affective wellbeing (happiness) between the rich and the poor
- Conversely the linear relationship between income and cognitive wellbeing (life-satisfaction) has remained constant over this period
- The happiness of an increasing proportion of the population is more dependent on their financial security than ever this millennium

@@

Fundementally, a difference in slope between affective wellbeing and income for the rich and poor represents an unacknowledged source of inequality in the distribution of wellbeing (i.e., happiness) in the economy. For instance, the "satiety" point of USD75,000 in 2008 was substantially more than the US median income of USD52,000 in the same year, indicating that the affective wellbeing of the poorest majority of the US population was tied to marginal changes in income while the affective wellbeing of a richer minority was not. While much commentary has focused on the zero slope at high income levels reported by Kahneman & Diener ("satiety" point), any decrease in slope for the richer minority would also represent a relatively unequal distribution of happiness among the poor compared to the rich. Thus, even a non-zero changepoint indicates that income is tied to inequities in the distribution of affective wellbeing in the economy. Conversely, a more equitable distribution of wellbeing in the economy would correspond to a lower valued change point.

While studies agree that affective wellbeing tends to increase with income, the exact form of the functional relationship between affective wellbeing and income is contested: some studies show the presence of a zero-slope changepoint after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]; other studies have found that affective wellbeing is a constant (albiet decreasing) log-linear function of income [@killingsworth2021experienced; @sacks2012new; @twenge2020expanding]. Few studies have formally tested the functional form of the relationship, relying instead upon binning the data (i.e., over income deciles or quintiles) and comparing adjacent bins either visually or with null hypothesis tests which do not provide evidence for a changepoint. In this regard, arbitrary choices about bin width can have salient effects on the function form, which in turn will influence the inferences drawn. And no studies to our knowledge have ever tested the presence of a changepoint between wellbeing and income using standard modelling approaches. For this reason, a formal comparison between log-linear, linear and a changepoint function (piecewise-linear) in the relationship between affective wellbeing and income is warranted.

@@

We found the marginal effect of household income on affective- and cognitive-wellbeing was positive but quite different: Affective wellbeing increased rapidly up to a point after which higher levels of income were associated with less improvement, while cognitive wellbeing tended to increase linearly or constantly with income. Other studies have also reported that income has distinct effects on cognitive and affective wellbeing [@howell2008relation; @kahneman2010high], with a zero-slope change point in affective wellbeing after which increasing income has little further effect. In our case the changepoint did not precede a zero-slope, as the post-slope for each year was credibly above zero (Figure 2), but it did represent a substantial decrease in dependency between income and affective wellbeing for those higher income levels. Our novel finding was revealing the change point increased faster than both inflation and median household income between 2001 and 2019. For the first time we have shown there has been a temporal shift in the change point between income and affective wellbeing over the 19 year period, such that affective wellbeing has become more dependent on income for more people - especially the poor and middle-class.

@@

Other researchers have argued income has a satiety point on affective well-being, after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]. This is consistent with a change point after which the slope is zero. While we observed a change-point, the slope afterwards was small but credibly larger than zero so our results are not strictly consistent with a "satiety" point. Our results emphasize the slope over income is not constant. Other researchers have argued that affective wellbeing is a constant (albiet decreasing) log-linear function of income [@killingsworth2021experienced; @sacks2012newl] or a constant linear relationship [@twenge2020expanding]. We also show the change-point has increased over time. Indeed Twenge et al report the linear relationship between income and happiness has become steeper over time in the US General Social Survey (GSS). While we did not see any consistent trend over time in pre or post-slope changes in the piecewise-linear model of affective-wellbeing (e.g., Figure 2), there was a credibly increasing slope in the linear model of cognitive-wellbeing with time (Table S3). It may be the single-item on general happiness in the GSS has more in common with the single item on life-satisfaction in HILDA than the more comprehensive 9-item, dual-component questions on affective-wellbeing in the SF-36. Nevertheless, like Twenge et al, we report that the difference in happiness betweeen the rich and poor has been increasing over time (e.g., Figure S6), however our piecewise-model reveals the income level around which this difference occurs.

>>>>

- We investigate how the relationship between income and subjective wellbeing (affective and cognitive) has changed over time, using a dataset with highly detailed measurements of each
- Nineteen years of data from a well-developed, economically stable G20 nation is examined (Australia), allowing a relatively undisturbed view of the changes without major economic shocks
- We are able to show for the first time that the income level at which affective wellbeing (happiness) is less determined by income has risen since 2001, resulting in greater inequality in happiness between the rich and the poor
- Conversely the linear relationship between income and cognitive wellbeing (life-satisfaction) has remained relatively constant over this period
- The happiness of an increasing proportion of the population is more dependent on their financial security than ever this millennium

@@

Fundementally, a difference in slope between affective wellbeing and income for the rich and poor represents an unacknowledged source of inequality in the distribution of wellbeing (i.e., happiness) in the economy. For instance, the "satiety" point of USD75,000 in 2008 was substantially more than the US median income of USD52,000 in the same year, indicating that the affective wellbeing of the poorest majority of the US population was tied to marginal changes in income while the affective wellbeing of a richer minority was not. While much commentary has focused on the zero slope at high income levels reported by Kahneman & Deaton ("satiety" point), any decrease in slope for the richer minority would also represent a relatively unequal distribution of happiness among the poor compared to the rich. Thus, even a non-zero changepoint indicates that income is tied to inequities in the distribution of affective wellbeing in the economy. Conversely, a more equitable distribution of wellbeing in the economy would correspond to a lower valued change point.

While studies agree that affective wellbeing tends to increase with income, the exact form of the functional relationship between affective wellbeing and income is contested: some studies show the presence of a zero-slope changepoint after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]; other studies have found that affective wellbeing is a constant (albiet decreasing) log-linear function of income [@killingsworth2021experienced; @sacks2012new; @twenge2020expanding]. Few studies have formally tested the functional form of the relationship, relying instead upon binning the data (i.e., over income deciles or quintiles) and comparing adjacent bins either visually or with null hypothesis tests which do not provide evidence for a changepoint. In this regard, arbitrary choices about bin width can have salient effects on the function form, which in turn will influence the inferences drawn. And no studies to our knowledge have ever tested the presence of a changepoint between wellbeing and income using standard modelling approaches. For this reason, a formal comparison between log-linear, linear and a changepoint function (i.e., piecewise) in the relationship between affective wellbeing and income is warranted.

@@

We found the marginal effect of household income on affective- and cognitive-wellbeing was positive but quite different: Affective wellbeing increased rapidly up to a point after which higher levels of income were associated with less improvement, while cognitive wellbeing tended to increase linearly or constantly with income. Other studies have also reported that income has distinct effects on cognitive and affective wellbeing [@howell2008relation], with a zero-slope change point in affective wellbeing after which increasing income has little further effect [@kahneman2010high]. In our case we did not observe a zero-slope changepoint ("satiety" point), as the post-slope for each year was credibly above zero (Figure 2); but the changepoint did represent a substantial decrease in dependency between income and affective wellbeing for those higher income levels. Our novel finding was revealing the changepoint increased faster than both inflation and median household income between 2001 and 2019. For the first time we have shown there has been a temporal shift in the changepoint between income and affective wellbeing over the 19 year period, such that affective wellbeing has become more dependent on income for more people - especially the poor and middle-class.

@@

Other researchers have argued income has a satiety point on affective well-being, after which increasing income has no further effect [@kahneman2010high; @jebb2018happiness]. This is consistent with a change point after which the slope is zero. While we observed a change-point, the slope afterwards was small but credibly larger than zero so our results are not strictly consistent with a "satiety" point. Our results emphasize the slope over income is not constant. Other researchers have argued that affective wellbeing is a constant (albiet decreasing) log-linear function of income [@killingsworth2021experienced; @sacks2012newl; @twenge2020expanding]. We also show the change-point has increased over time. Indeed Twenge & Cooper report the linear relationship between income and happiness has become steeper over time in the US General Social Survey (GSS). While we did not see any consistent trend over time in pre or post-slope changes in the piecewise-linear model of affective-wellbeing (e.g., Figure 2), there was a credibly increasing slope in the linear model of cognitive-wellbeing with time (Table S3). It may be the single-item on general happiness in the GSS has more in common with the single item on life-satisfaction in HILDA than the more comprehensive 9-item, dual-valence questions on affective-wellbeing in the SF-36. Nevertheless, like Twenge et al, we report that the difference in happiness betweeen the rich and poor has been increasing over time (e.g., Figure S6), moreover our piecewise-model reveals the income level around which this difference appears.
